% LaTeX source for textbook ``Physical Modeling in MATLAB''
% Copyright 2012, 2019, 2021 Allen B. Downey

\documentclass{book}

\newcommand{\thetitle}{Physical Modeling in MATLAB}
\newcommand{\theauthors}{Allen B. Downey}
\newcommand{\theversion}{4.0}

\title{\thetitle}
\author{\theauthors}
\date{Version \theversion}

\input{settings}

\makeindex

%\makeatletter
%\def\idxdelim{\@ifnextchar{\hyperindexformat}{. }{, }}
%\makeatother

\begin{document}
\sloppy

\frontmatter




\maketitle

\vspace{2in}

\begin{center}
 {\Large \thetitle}

\vspace{0.25in}

Copyright 2012, 2019 Allen B. Downey
\end{center}

\vspace{0.25in}

\begin{flushleft}
Green Tea Press       \\
9 Washburn Ave \\
Needham MA 02492
\end{flushleft}

Permission is granted to copy, distribute, and/or modify this document under the terms of the Creative Commons Attribution-NonCommercial 4.0 Unported License, which is available at \url{https://greenteapress.com/matlab/license}.

This book was typeset by the author using pdflatex, among other free, open-source programs.
 The LaTeX source for this book is available from \url{https://greenteapress.com/matlab}.

MATLAB\myreg\ is a registered trademark of The Mathworks, Inc.  The Mathworks does not warrant the accuracy  of this book.

The information in this book is distributed on an ``As Is'' basis,
without warranty. While every precaution has been taken in the
preparation of this work, neither the author nor No Starch Press, Inc.
shall have any liability to any person or entity with respect to any
loss or damage caused or alleged to be caused directly or indirectly
by the information contained in it.


ISBN-10: 1-23456-789-0\\
ISBN-13: 123-4-56789-012-3


\begin{verbatim}
Uthor, A. U. and Respondent, C. O. R. and Iter, W. R.
  Pellentesque habitant morbi: Cum sociis natoque penatibus/
  A. U. Thor, C. O. R. Respondent and W. R. Iter
       p. cm.
  Includes index.
  ISBN-10: 1-23456-789-0
  ISBN-13: 123-4-56789-012-3
  1.  Morbi (habitant).  2. Penatibus. I. Title.
XX2303.5.T324 2008
\end{verbatim}




\thispagestyle{empty}

\null\vskip1.175in

\section*{About the Author}
Allen Downey is an author and professor with more than 20 years of teaching experience at the college level. He helped create Olin College, a new engineering school with the mission to create an innovate curriculum, and to transform engineering education. He has also written more than 10 books about programming and using computation to learn about topics in math, science, and engineering.

\tableofcontents

\include{chap00}

\mainmatter


\include{chap01}

\include{chap02}

\include{chap03}

\include{chap04}

\include{chap05}

\include{chap06}

\include{chap07}

\include{chap08}

\include{chap09}

\include{chap10}

\include{chap11}

\include{chap12}

\include{chap13}

\include{chap14}

\include{chap15}

\include{glossary}

\printindex

\end{document}
